# R and Pandas Dataframe Comparison

## Introduction

Two of the most powerful tools for data analysis are R dataframes and Python pandas dataframes. While they serve similar purposes, they have different syntaxes, philosophies, and strengths. This chapter provides hands-on exercises to help you understand both approaches and make informed decisions about which to use for different tasks.

## Why Compare R and Pandas?

- **R** has deep statistical roots and excels at statistical analysis and visualization
- **Pandas** brings powerful data manipulation to Python's broader ecosystem
- Many research teams use both, and data often needs to move between them
- Understanding both helps you choose the right tool for each project
- The concepts transfer: learning one makes learning the other easier

## Prerequisites

- You have R installed and can run R code
- You have Python installed with pandas library (`pip install pandas` or `uv pip install pandas`)
- You can execute code in either RStudio, Jupyter notebooks, or your preferred IDE

```{r}
#| echo: false
library(reticulate)
```

## Dataframe Self-Learning Exercises: R and Python Pandas

---

## Part 1: R Dataframe Exercises

### Exercise Set R1: Creating and Exploring Dataframes

#### R1.1 Create a Research Dataset
**Task:** Create a dataframe with experimental data.

```{r}
#| eval: false

# Create a dataframe with participant data
participants <- data.frame(
  subject_id = c(101, 102, 103, 104, 105, 106, 107, 108, 109, 110),
  age = c(22, 25, 23, 28, 21, 24, 26, 23, 27, 22),
  condition = c("control", "treatment", "control", "treatment", "control", 
                "treatment", "control", "treatment", "control", "treatment"),
  reaction_time = c(450, 380, 470, 360, 490, 370, 460, 350, 480, 365),
  accuracy = c(0.85, 0.92, 0.83, 0.95, 0.80, 0.93, 0.84, 0.96, 0.82, 0.94)
)

# View the dataframe
print(participants)

# Check structure
str(participants)
```

**Questions to answer:** 
- How many rows and columns does your dataframe have?
- What data types are in each column?
- Is `condition` stored as a character or factor?

<details>
<summary>Expected outcome</summary>

You should see a dataframe with 10 rows and 5 columns. The `str()` function shows that `condition` is stored as a character vector by default in modern R. You can see the data types: numeric for age, reaction_time, and accuracy; character for condition; numeric for subject_id.
</details>

#### R1.2 Basic Dataframe Information
**Task:** Explore your dataframe's properties.

```{r}
#| eval: false

# Number of rows
nrow(participants)

# Number of columns
ncol(participants)

# Dimensions (rows, columns)
dim(participants)

# Column names
names(participants)
# or
colnames(participants)

# First few rows
head(participants)

# Last few rows
tail(participants)

# Summary statistics
summary(participants)
```

**Questions to answer:**
- What does `summary()` tell you about each column?
- How does it handle numeric vs. character columns differently?

<details>
<summary>Expected outcome</summary>

`summary()` provides descriptive statistics (min, max, mean, median, quartiles) for numeric columns and counts for character/factor columns. This is a quick way to check your data for outliers or data entry errors.
</details>

#### R1.3 Selecting Columns
**Task:** Practice different ways to access columns.

```{r}
#| eval: false

# Method 1: Dollar sign notation
participants$reaction_time

# Method 2: Square brackets with column name
participants[, "reaction_time"]

# Method 3: Square brackets with column number
participants[, 4]

# Select multiple columns
participants[, c("subject_id", "reaction_time", "accuracy")]

# Using subset to select columns
subset(participants, select = c(subject_id, reaction_time))
```

**Questions to answer:**
- Which method do you find most readable?
- What's the difference between `participants$reaction_time` and `participants[, "reaction_time"]`?
- What happens if you use single brackets: `participants["reaction_time"]`?

<details>
<summary>Expected outcome</summary>

`$` returns a vector, while `[, "column"]` returns a vector by default but can return a dataframe if you use `drop=FALSE`. Single brackets `["column"]` returns a dataframe with one column. Understanding these differences is important for avoiding errors in your analysis code.
</details>

#### R1.4 Filtering Rows
**Task:** Select subsets of your data based on conditions.

```{r}
#| eval: false

# Filter for treatment condition only
treatment_group <- participants[participants$condition == "treatment", ]
print(treatment_group)

# Filter for participants younger than 25
young_participants <- participants[participants$age < 25, ]
print(young_participants)

# Multiple conditions: treatment AND age < 25
young_treatment <- participants[participants$condition == "treatment" & 
                                participants$age < 25, ]
print(young_treatment)

# Using subset function (often more readable)
treatment_group2 <- subset(participants, condition == "treatment")
young_treatment2 <- subset(participants, condition == "treatment" & age < 25)
```

**Questions to answer:**
- How many participants are in the treatment group?
- How many young participants (age < 25) received treatment?
- Which syntax do you find clearer: bracket notation or `subset()`?

<details>
<summary>Expected outcome</summary>

You should find 5 participants in treatment group, and 2 young participants in treatment. The `subset()` function is often more readable because you don't need to repeat the dataframe name and the comma notation is less error-prone.
</details>

#### R1.5 Computing Condition Means
**Task:** Calculate mean reaction time by condition.

```{r}
#| eval: false

# Method 1: Manual filtering
control_rt <- participants[participants$condition == "control", "reaction_time"]
treatment_rt <- participants[participants$condition == "treatment", "reaction_time"]

mean(control_rt)
mean(treatment_rt)

# Method 2: Using aggregate
aggregate(reaction_time ~ condition, data = participants, FUN = mean)

# Method 3: Using tapply
tapply(participants$reaction_time, participants$condition, mean)

# Get both mean and median
aggregate(reaction_time ~ condition, data = participants, 
          FUN = function(x) c(mean = mean(x), median = median(x)))
```

**Questions to answer:**
- Which condition has faster reaction times on average?
- Which method do you find most intuitive?
- What's the difference between the output formats of `aggregate()` and `tapply()`?

<details>
<summary>Expected outcome</summary>

Treatment group should have faster reaction times (lower values). `aggregate()` returns a dataframe which is often easier to work with, while `tapply()` returns a named vector. The formula syntax in `aggregate()` (`reaction_time ~ condition`) is very readable and common in R statistical functions.
</details>

#### R1.6 Adding and Modifying Columns
**Task:** Create new variables in your dataframe.

```{r}
#| eval: false

# Add a new column: convert RT from ms to seconds
participants$rt_seconds <- participants$reaction_time / 1000

# Create a categorical age group
participants$age_group <- ifelse(participants$age < 25, "young", "older")

# Create a performance score (combining RT and accuracy)
# Lower RT and higher accuracy = better performance
participants$performance <- participants$accuracy / (participants$reaction_time / 1000)

# View the updated dataframe
head(participants)

# Check the new columns
summary(participants)
```

**Questions to answer:**
- How many columns does your dataframe have now?
- What are the mean performance scores for each condition?
- How would you create a three-level age group (young, middle, older)?

<details>
<summary>Expected outcome</summary>

Your dataframe now has 8 columns. You can calculate condition means for performance using the methods from R1.5. For three-level grouping, you'd use nested `ifelse()` or the `cut()` function.
</details>

#### R1.7 Basic Statistical Tests
**Task:** Perform a t-test comparing conditions.

```{r}
#| eval: false

# T-test: Does treatment affect reaction time?
t.test(reaction_time ~ condition, data = participants)

# Alternative: manual specification
control_rt <- participants$reaction_time[participants$condition == "control"]
treatment_rt <- participants$reaction_time[participants$condition == "treatment"]
t.test(control_rt, treatment_rt)

# Correlation: Age and reaction time
cor.test(participants$age, participants$reaction_time)

# Correlation: Reaction time and accuracy
cor.test(participants$reaction_time, participants$accuracy)
```

**Questions to answer:**
- Is there a significant difference in reaction time between conditions?
- What is the correlation between reaction time and accuracy?
- What does a negative correlation between RT and accuracy suggest?

<details>
<summary>Expected outcome</summary>

You should see a significant difference between conditions (treatment faster). The correlation between RT and accuracy should be negative (faster responses = higher accuracy in this dataset), suggesting a speed-accuracy tradeoff is NOT present - participants who are faster are also more accurate.
</details>

---

## Part 2: Python Pandas Dataframe Exercises

### Exercise Set P1: Creating and Exploring Dataframes

#### P1.1 Create the Same Research Dataset
**Task:** Create an identical dataframe in pandas.

```{python}
#| eval: false

import pandas as pd
import numpy as np

# Create a dataframe with participant data
participants = pd.DataFrame({
    'subject_id': [101, 102, 103, 104, 105, 106, 107, 108, 109, 110],
    'age': [22, 25, 23, 28, 21, 24, 26, 23, 27, 22],
    'condition': ['control', 'treatment', 'control', 'treatment', 'control',
                  'treatment', 'control', 'treatment', 'control', 'treatment'],
    'reaction_time': [450, 380, 470, 360, 490, 370, 460, 350, 480, 365],
    'accuracy': [0.85, 0.92, 0.83, 0.95, 0.80, 0.93, 0.84, 0.96, 0.82, 0.94]
})

# View the dataframe
print(participants)

# Check structure
print(participants.info())
print(participants.dtypes)
```

**Questions to answer:**
- How does the pandas output compare to R's `print()`?
- What data types does pandas assign to each column?
- How is `info()` similar to or different from R's `str()`?

<details>
<summary>Expected outcome</summary>

Pandas displays the dataframe with row indices (0-9) on the left. The `info()` method shows similar information to R's `str()`: column names, non-null counts, and data types. Pandas uses `int64`, `float64`, and `object` (for strings) as data types.
</details>

#### P1.2 Basic Dataframe Information
**Task:** Explore your dataframe's properties.

```{python}
#| eval: false

# Number of rows and columns
print(participants.shape)  # Returns (rows, columns)

# Number of rows
print(len(participants))

# Column names
print(participants.columns)

# First few rows
print(participants.head())

# Last few rows
print(participants.tail())

# Summary statistics
print(participants.describe())

# More detailed summary including non-numeric columns
print(participants.describe(include='all'))
```

**Questions to answer:**
- How does `shape` compare to R's `dim()`?
- What does `describe()` show you? How is it different from R's `summary()`?
- What happens when you use `describe(include='all')`?

<details>
<summary>Expected outcome</summary>

`shape` is a tuple (10, 5) similar to R's `dim()`. `describe()` shows statistics for numeric columns only by default, while R's `summary()` includes all columns. Using `include='all'` shows statistics for all column types, including count, unique values, and most frequent value for categorical data.
</details>

#### P1.3 Selecting Columns
**Task:** Practice different ways to access columns.

```{python}
#| eval: false

# Method 1: Bracket notation with column name
print(participants['reaction_time'])

# Method 2: Dot notation (only works if column name has no spaces/special chars)
print(participants.reaction_time)

# Select multiple columns - returns a dataframe
print(participants[['subject_id', 'reaction_time', 'accuracy']])

# Using loc for label-based selection
print(participants.loc[:, 'reaction_time'])

# Using iloc for position-based selection
print(participants.iloc[:, 3])  # 4th column (0-indexed)
```

**Questions to answer:**
- What's the difference between `participants['reaction_time']` and `participants[['reaction_time']]`?
- When would you use `loc` vs `iloc`?
- How does pandas column selection compare to R's methods?

<details>
<summary>Expected outcome</summary>

Single brackets with one column name returns a Series (like a vector), double brackets returns a DataFrame. `loc` uses labels (column/row names), `iloc` uses integer positions. This is similar to R's distinction between `$` (returns vector) and `[, "col"]` (can return vector or dataframe).
</details>

#### P1.4 Filtering Rows
**Task:** Select subsets of your data based on conditions.

```{python}
#| eval: false

# Filter for treatment condition only
treatment_group = participants[participants['condition'] == 'treatment']
print(treatment_group)

# Filter for participants younger than 25
young_participants = participants[participants['age'] < 25]
print(young_participants)

# Multiple conditions: treatment AND age < 25
young_treatment = participants[(participants['condition'] == 'treatment') & 
                               (participants['age'] < 25)]
print(young_treatment)

# Using query method (often more readable)
treatment_group2 = participants.query('condition == "treatment"')
young_treatment2 = participants.query('condition == "treatment" and age < 25')
print(young_treatment2)
```

**Questions to answer:**
- Why do you need parentheses around each condition when using `&`?
- How many participants are in the treatment group?
- How does `query()` compare to R's `subset()`?

<details>
<summary>Expected outcome</summary>

Parentheses are needed because of operator precedence in Python. You should find 5 treatment participants and 2 young treatment participants. The `query()` method is similar to R's `subset()` in readability, using string expressions instead of boolean indexing.
</details>

#### P1.5 Computing Condition Means
**Task:** Calculate mean reaction time by condition.

```{python}
#| eval: false

# Method 1: Manual filtering
control_rt = participants[participants['condition'] == 'control']['reaction_time']
treatment_rt = participants[participants['condition'] == 'treatment']['reaction_time']

print(f"Control mean: {control_rt.mean()}")
print(f"Treatment mean: {treatment_rt.mean()}")

# Method 2: Using groupby
print(participants.groupby('condition')['reaction_time'].mean())

# Method 3: Using groupby with agg for multiple statistics
print(participants.groupby('condition')['reaction_time'].agg(['mean', 'median', 'std']))

# Multiple columns at once
print(participants.groupby('condition')[['reaction_time', 'accuracy']].mean())
```

**Questions to answer:**
- Which condition has faster reaction times?
- How does `groupby()` compare to R's `aggregate()` or `tapply()`?
- What does the `agg()` method allow you to do?

<details>
<summary>Expected outcome</summary>

Treatment group has faster reaction times. `groupby()` is similar to R's `aggregate()` but often more flexible. The `agg()` method lets you apply multiple functions at once, returning a nice summary table. This is very powerful for exploratory data analysis.
</details>

#### P1.6 Adding and Modifying Columns
**Task:** Create new variables in your dataframe.

```{python}
#| eval: false

# Add a new column: convert RT from ms to seconds
participants['rt_seconds'] = participants['reaction_time'] / 1000

# Create a categorical age group
participants['age_group'] = participants['age'].apply(
    lambda x: 'young' if x < 25 else 'older'
)

# Alternative using numpy where (similar to ifelse)
participants['age_group2'] = np.where(participants['age'] < 25, 'young', 'older')

# Create a performance score
participants['performance'] = participants['accuracy'] / (participants['reaction_time'] / 1000)

# View the updated dataframe
print(participants.head())

# Check the new columns
print(participants.describe())
```

**Questions to answer:**
- How many columns does your dataframe have now?
- What's the difference between using `apply()` with a lambda vs `np.where()`?
- How does adding columns in pandas compare to R?

<details>
<summary>Expected outcome</summary>

Your dataframe now has 9 columns (one more than R because we created age_group twice). `apply()` with lambda is more flexible but can be slower; `np.where()` is vectorized and faster, similar to R's `ifelse()`. Both languages make it easy to add columns with simple assignment.
</details>

#### P1.7 Basic Statistical Tests
**Task:** Perform a t-test comparing conditions.

```{python}
#| eval: false

from scipy import stats

# T-test: Does treatment affect reaction time?
control_rt = participants[participants['condition'] == 'control']['reaction_time']
treatment_rt = participants[participants['condition'] == 'treatment']['reaction_time']

t_stat, p_value = stats.ttest_ind(control_rt, treatment_rt)
print(f"T-test results: t={t_stat:.3f}, p={p_value:.4f}")

# Correlation: Age and reaction time
corr_age_rt, p_age_rt = stats.pearsonr(participants['age'], participants['reaction_time'])
print(f"Correlation (age, RT): r={corr_age_rt:.3f}, p={p_age_rt:.4f}")

# Correlation: Reaction time and accuracy
corr_rt_acc, p_rt_acc = stats.pearsonr(participants['reaction_time'], participants['accuracy'])
print(f"Correlation (RT, accuracy): r={corr_rt_acc:.3f}, p={p_rt_acc:.4f}")

# Using pandas corr() for correlation matrix
print(participants[['age', 'reaction_time', 'accuracy']].corr())
```

**Questions to answer:**
- Are the statistical results the same as in R?
- How does scipy's output format differ from R's?
- What does the correlation matrix show you?

<details>
<summary>Expected outcome</summary>

Results should be identical to R (within rounding). Scipy returns tuples (statistic, p-value) rather than R's detailed output objects. The correlation matrix shows all pairwise correlations at once, which is useful for exploring relationships in your data.
</details>

---

## Part 3: Comparative Exercises

### Exercise C1: Side-by-Side Comparison
**Task:** Work through the same analysis in both languages and compare.

**Scenario:** You want to know if there's a relationship between age and performance, separately for each condition.

**R Version:**
```{r}
#| eval: false

# Calculate correlation for each condition
by(participants[, c("age", "performance")], 
   participants$condition, 
   function(x) cor(x$age, x$performance))

# Or using aggregate with a custom function
aggregate(cbind(age, performance) ~ condition, 
          data = participants,
          FUN = function(x) cor(x[,1], x[,2]))
```

**Python Version:**
```{python}
#| eval: false

# Calculate correlation for each condition
participants.groupby('condition').apply(
    lambda x: x['age'].corr(x['performance'])
)

# Alternative: manual calculation
for condition in participants['condition'].unique():
    subset = participants[participants['condition'] == condition]
    corr = subset['age'].corr(subset['performance'])
    print(f"{condition}: {corr:.3f}")
```

**Questions to answer:**
- Which approach feels more natural to you?
- How do the syntax patterns differ?
- Which language makes this task easier?

<details>
<summary>Expected outcome</summary>

Both accomplish the same task but with different syntax. R's formula interface (`~ condition`) is very concise. Python's `groupby().apply()` is explicit about the grouping and function application. Neither is objectively "better" - it depends on your background and preferences.
</details>

### Exercise C2: Data Reshaping Challenge
**Task:** Create a summary table showing mean RT and accuracy by condition.

**R Version:**
```{r}
#| eval: false

# Using aggregate
summary_table <- aggregate(cbind(reaction_time, accuracy) ~ condition,
                          data = participants,
                          FUN = mean)
print(summary_table)

# Using dplyr (if available)
# library(dplyr)
# participants %>%
#   group_by(condition) %>%
#   summarise(mean_rt = mean(reaction_time),
#            mean_acc = mean(accuracy))
```

**Python Version:**
```{python}
#| eval: false

# Using groupby with agg
summary_table = participants.groupby('condition')[['reaction_time', 'accuracy']].mean()
print(summary_table)

# Reset index to make condition a column instead of index
summary_table_reset = participants.groupby('condition')[['reaction_time', 'accuracy']].mean().reset_index()
print(summary_table_reset)

# Rename columns for clarity
summary_table_renamed = participants.groupby('condition').agg({
    'reaction_time': 'mean',
    'accuracy': 'mean'
}).rename(columns={'reaction_time': 'mean_rt', 'accuracy': 'mean_acc'}).reset_index()
print(summary_table_renamed)
```

**Questions to answer:**
- How do the default outputs differ between R and pandas?
- What is the pandas "index" and how does it compare to R's row names?
- When would you use `reset_index()`?

<details>
<summary>Expected outcome</summary>

R returns a dataframe with condition as a regular column. Pandas makes condition the index by default. The index is similar to R's row names but more powerful. Use `reset_index()` when you want the grouping variable as a regular column for further analysis or export.
</details>

### Exercise C3: Saving Your Work
**Task:** Save your dataframe to a CSV file in both languages.

**R Version:**
```{r}
#| eval: false

# Save to CSV
write.csv(participants, "participants_r.csv", row.names = FALSE)

# Read it back
participants_loaded <- read.csv("participants_r.csv")

# Check if it's the same
identical(participants, participants_loaded)
```

**Python Version:**
```{python}
#| eval: false

# Save to CSV
participants.to_csv("participants_python.csv", index=False)

# Read it back
participants_loaded = pd.read_csv("participants_python.csv")

# Check if it's the same
print(participants.equals(participants_loaded))

# Check what's different if not equal
print(participants.dtypes)
print(participants_loaded.dtypes)
```

**Questions to answer:**
- Why do we use `row.names = FALSE` in R and `index=False` in pandas?
- Are the loaded dataframes identical to the originals?
- What happens to data types when you save and reload?

<details>
<summary>Expected outcome</summary>

Both languages can save/load CSV files easily. Setting `row.names/index = FALSE` prevents writing row numbers to the file. Data types might change slightly (e.g., integers might become floats) depending on the CSV format, but the data values should be the same.
</details>

---

## Reflection Questions

After completing all exercises, consider:

1. **Syntax Patterns:**
   - R uses `$` for column access; pandas uses `[]` or `.`
   - R uses formula notation (`y ~ x`); pandas uses method chaining
   - Which feels more intuitive to you?

2. **Indexing:**
   - R uses 1-based indexing; Python uses 0-based indexing
   - R has row names; pandas has a more powerful index system
   - How does this affect your code?

3. **Function vs. Method:**
   - R: `mean(df$column)` - function applied to vector
   - Pandas: `df['column'].mean()` - method called on object
   - What are the advantages of each approach?

4. **Grouping Operations:**
   - R: `aggregate()`, `tapply()`, `by()`
   - Pandas: `groupby()` with method chaining
   - Which is more flexible? Which is more readable?

5. **Statistical Functions:**
   - R has statistics built into the base language
   - Python requires importing scipy
   - Does this matter for your workflow?

---

## Summary: Key Similarities and Differences

### Similarities
- Both have dataframe structures with rows and columns
- Both can filter, select, and transform data
- Both can compute grouped statistics
- Both can save/load common file formats
- Both have extensive statistical capabilities

### Key Differences

| Aspect | R | Pandas |
|--------|---|--------|
| Column access | `df$col` or `df[, "col"]` | `df['col']` or `df.col` |
| Multiple columns | `df[, c("a", "b")]` | `df[['a', 'b']]` |
| Filtering | `df[df$x > 5, ]` | `df[df['x'] > 5]` |
| Grouping | `aggregate(y ~ group, df, mean)` | `df.groupby('group')['y'].mean()` |
| Adding columns | `df$new <- value` | `df['new'] = value` |
| Method vs function | Functions: `mean(df$x)` | Methods: `df['x'].mean()` |
| Indexing | 1-based | 0-based |
| Missing values | `NA` | `NaN` or `None` |

---

## Next Steps

Once comfortable with these basics, explore:

**R:**
- `dplyr` package for data manipulation (very popular)
- `tidyr` for reshaping data
- `ggplot2` for visualization
- `data.table` for large datasets

**Python Pandas:**
- Advanced groupby operations
- Merging and joining dataframes
- Time series functionality
- Integration with scikit-learn for machine learning

**Both:**
- Handling missing data
- Working with dates and times
- Merging datasets
- Pivot tables and cross-tabulations

---

## Common Pitfalls

### R
- Forgetting the comma in `df[condition, ]`
- Factor vs. character confusion
- Recycling rules can cause unexpected behavior

### Pandas
- Forgetting that many operations return a copy, not modifying in place
- Index alignment can cause unexpected results
- Chained indexing warnings (`df[df['x'] > 5]['y'] = 10` is problematic)

### Both
- Not checking data types after loading from files
- Assuming missing values are handled the same way
- Not verifying results with small test cases first

---

**License:** This document is released under CC BY 4.0. Feel free to adapt for your courses.





