## An Example Lab Notebook {#sec-lab-notebook-example}

```yaml
---
# 1. METADATA: THE OFFICIAL RECORD 
# (Purpose: This section contains essential information for identifying, tracking, and citing the research activity. It mimics the cover page or mandatory headers of a bound notebook.)
title: "PSYC [Course Number] Lab Notebook Entry: [Insert Activity Title]"
author: "Student Name(s) and ID(s)"
date: today # Automatically uses the current date. This serves as the official timestamp for the entry.
# output: 
#   html: 
#     toc: true # Creates a Table of Contents for easy navigation
#     embed-resources: true # Makes the final HTML file self-contained
#   pdf: default # Uncomment if you prefer a PDF output
---
```



## 🎯 1. Research Aim & Hypotheses

### Research Question
What is the effect of **[Independent Variable (IV)]** on **[Dependent Variable (DV)]** in our sample of [Population]?

### Hypotheses
* **H1 (Alternative):** We predict that [Specific prediction based on IV and DV].
* **H0 (Null):** There will be no significant effect of the IV on the DV.

## 🔬 2. Methodology & Procedure

### Participants
* **Recruitment Method:** [e.g., SONA, convenience sample]
* **Target Sample Size:** [Number]
* **Demographic Criteria:** [e.g., Age 18-25, native English speakers]

### Materials & Apparatus
* **Psychological Measure/Scale:** [e.g., State-Trait Anxiety Inventory (STAI)]
* **Stimuli Details:** [e.g., 20 high-arousal images from the IAPS database]
* **Software/Equipment:** [e.g., PsychoPy, Qualtrics, R Studio]

### Step-by-Step Procedure
1.  [Detailed action 1, e.g., Calibrated the response box.]
2.  [Detailed action 2, e.g., Participant received standardized instructions.]
3.  [Detailed action 3, e.g., Ran 4 blocks of 20 trials each, with a 30-second break between blocks.]

### 🚨 Critical Observations & Deviations (The "Activity Log")

> **Observation 1 (Time: 10:45 AM):** The internet connection briefly dropped for 3 minutes during the survey administration for Participant 7. The system appears to have saved their progress, but a note was made to check the final file for completeness.
>
> **Deviation:** We were supposed to collect 20 participants, but due to time, only collected 18. Analysis must account for lower statistical power.

## 📊 3. Data Processing & Analysis

### 3.01 Generating Fake Data
This is **not** part of a lab notebook. This is some code to generate random data so that the rest of the code can be tested. 

```{r}
# Lab Notebook Data Generation Script
# Purpose: Generates a sample CSV file with clean, pre-aggregated data
#          for use in the psychology lab notebook lesson (Quarto/R).

# Load necessary library
library(tidyverse)

# --- Define Parameters ---
N_PARTICIPANTS <- 25
CONDITIONS <- c("Treatment", "Control")

# Parameters for Accuracy (The Dependent Variable)
MEAN_TREATMENT_ACC <- 0.85 
MEAN_CONTROL_ACC <- 0.75 
SD_ACCURACY <- 0.10

# Parameters for Reaction Time (The Filtering Variable)
MEAN_RT <- 650 # in milliseconds
SD_RT <- 150 # in milliseconds

# --- Generate the Data ---
set.seed(42) # Set a seed for reproducible random data

sample_raw_data <- tibble(
  participant_id = 1:N_PARTICIPANTS,
  condition = factor(rep(CONDITIONS, length.out = N_PARTICIPANTS)),
  
  # Filtering Column: 90% pass (1), 10% fail (0).
  attention_check = rbinom(N_PARTICIPANTS, 1, 0.9), 
  
  # 1. ACCURACY Column (Dependent Variable, continuous score 0.0 to 1.0)
  # This column simulates the final aggregated accuracy score.
  accuracy = if_else(
    condition == "Treatment",
    rnorm(N_PARTICIPANTS, mean = MEAN_TREATMENT_ACC, sd = SD_ACCURACY),
    rnorm(N_PARTICIPANTS, mean = MEAN_CONTROL_ACC, sd = SD_ACCURACY)
  ) |> pmin(1.0) |> pmax(0.0), # Cap values between 0 and 1
  
  # 2. RT Column (Filtering Variable, in milliseconds)
  # This column simulates the mean reaction time for filtering.
  rt = round(rnorm(N_PARTICIPANTS, mean = MEAN_RT, sd = SD_RT))
)

# Introduce simulated outliers to test the filtering step in the notebook
# Participant 5 is too slow (tests RT filter)
sample_raw_data$rt[5] <- 3200 
# Participant 10 fails attention check and is too slow (tests both filters)
sample_raw_data$rt[10] <- 4500 
sample_raw_data$attention_check[10] <- 0 

# --- Save the Data ---
# Create the 'data' directory if it doesn't exist (matching the template's path)
if (!dir.exists("data")) {
  dir.create("data")
}

# Save the dataset as a CSV file
write_csv(sample_raw_data, "data/psych_exp_raw_data_2025-11-01.csv")

# Final check of the data structure
print("Final sample data file created successfully in the 'data' folder.")
print(head(sample_raw_data))
```

### 3.1 Setup and Data Import

```{r}
#| label: setup-and-import
#| echo: true # Value: Always show the code for importing data to confirm the source file.

# Load necessary packages (e.g., for data manipulation, statistics)

raw_data <- read_csv("data/psych_exp_raw_data_2025-11-01.csv") 

# Display the dimensions to check if import was successful
dim(raw_data)

```

```{r}
#| label: data-cleaning
#| echo: true 

# Filter out participants who failed an attention check (Reaction Time > 3000ms)
cleaned_data <- raw_data |>
  filter(rt < 3000) |>
  # Calculate the primary dependent variable (mean accuracy across all trials)
  mutate(mean_accuracy = mean(accuracy))

# Value: Always check the final sample size after filtering.
n_final <- nrow(cleaned_data)
cat("Final sample size used for analysis after exclusions: ", n_final, "\n")
```


### Example: Run a two-sample t-test to compare mean accuracy between two conditions (Treatment vs. Control)
```{r}
#| label: main-analysis
#| echo: true


t_test_result <- t.test(accuracy ~ condition, data = cleaned_data)

# Print the full results for comprehensive record-keeping
t_test_result
```

### Example: Generate a plot of the primary finding

```{r}
#| label: main-figure
#| echo: true
#| fig-cap: "Mean Accuracy Scores by Condition (Treatment vs. Control)"

cleaned_data |>
  group_by(condition) |>
  summarize(mean_acc = mean(accuracy), se = sd(accuracy)/sqrt(n())) |>
  ggplot(aes(x = condition, y = mean_acc)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_errorbar(aes(ymin = mean_acc - se, ymax = mean_acc + se), width = 0.2) +
  labs(y = "Mean Accuracy", x = "Experimental Condition") +
  theme_minimal()
```

## 🗓️ 4. Project Chronology and Daily Log

*(Purpose: This section serves as the **scientific diary** for the project. Every day you work on the research—whether collecting data, cleaning code, or reading literature—you must log your activities, thoughts, and any problems encountered. This maintains the chain of custody and the historical record of the research.)*

### 4.1 Log Entry: 2025-11-04 (Data Collection Session 2)

* **Time:** 1:00 PM - 4:00 PM.
* **Activity:** Collected data from 8 new participants (P19 - P26) using the same protocol as the first session.
* **Observations:** The data collection went smoothly, but **Participant 22** reported feeling tired and may have rushed the final block. No technical errors.
* **New Files Generated:** `raw_data_2025-11-04.csv`.
* **Action Required:** Merge `raw_data_2025-11-04.csv` with the existing dataset before the next analysis run.

### 4.2 Log Entry: 2025-11-05 (Literature Review & Code Cleanup)

* **Time:** 9:00 AM - 11:30 AM.
* **Activity:** Reviewed literature on the IV. Found a key paper suggesting the effect is stronger in female participants.
* **Thought/Reflection:** *I should add a gender variable to the final model to check for this interaction effect (a moderation analysis).*
* **Code Activity:** Renamed the `cleaned_data` object in the analysis script to `data_for_t_test` for better clarity.
* **Action Required:** Plan the code structure for running an **ANOVA** with gender as an additional factor next week.

### 4.3 Log Entry: 2025-11-10 (Initial Analysis Run & Troubleshooting)

* **Time:** 1:30 PM - 2:30 PM.
* **Activity:** Ran the `main-analysis` chunk in the Quarto notebook (Section 3.3) on the full dataset (P1 - P26).
* **Problem:** The `t.test()` code returned an error: `"Error: variable lengths differ."`
* **Resolution:** Identified that the error was caused by a participant's row being deleted in the **cleaning** step without the corresponding row being deleted in the original R object. The issue was resolved by reloading `raw_data` inside the `setup-and-import` chunk, ensuring the entire workflow is self-contained and reproducible.
* **Conclusion:** The p-value for the main effect is $p = 0.08$. The effect is trending, but not significant. More data is required.