# R

## Installation
See the operating system specific appendices.

### Test your installation

```{sh}
R cmd --version
```

You should see some information about you R version.

## Interfacing with VS Code
Restart VS Code and install the R extension (look to the left panel for the extension launcher; it looks like a little set of squares).

### Resources
[R for Data Science](https://r4ds.hadley.nz/) is a text book on data science that uses R. The author is a prolific R package author and the itself written in Quarto. It includes many examples that you can test by cutting and pasting into your IDE.

## Installing R Packages
There are many tools for this. Some of them are GUI tools, and they will at first appear easier. But if your library starts to grow, or you need to install particular versions of packages, old packages, or packages from Github and not on CRAN, then you will be much happier in a R interpreter.

To create an R interpreter in VS Code you can simply launch a new terminal and then type R and <enter>. You should see an R welcome message. Code like the snippet to follow shows a basic installation exchange. 

::: {#fig-installRpackage}

```r
install.packages("tidyverse",repos='https://mirror.csclub.uwaterloo.ca/CRAN/')
library{tidyverse}
```
:::

The [tidyverse](https://www.tidyverse.org/) is a very popular collection of R code that itself will depend on many additional pieces of R code and other packages, some R and some not.

## Mix R Code with Text for Reproducible and Documented Workflows 

Here is an example of embedding R code in a qmd document. In this code we download and install a package if necessary (the `if` statement). Then we use that package to load some data and generate a plot. The plot you see in this document is the result of that code. I did not cut and paste a png. 

```{r}
#| warning: false
if(!require(tidyverse)){
    install.packages("tidyverse",repos='https://mirror.csclub.uwaterloo.ca/CRAN/')
}
if(!require(palmerpenguins)){
    install.packages("palmerpenguins",repos='https://mirror.csclub.uwaterloo.ca/CRAN/')
}
library(tidyverse)
library(palmerpenguins)

ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point() +
  geom_smooth(method = "lm")
```

## R Types
As mentioned in the code chapter, programming language values are typically typed. R in particular has many complicated data types that represent the output of functions and statistical analyses. If things are getting confusing you might want to check on the type of your data.

::: {#fig-rchecktype}

```{r}
a = 1
typeof(a)
```

Checking the type of a number in R.

:::




::: {#fig-rdatatypes}

```{r}
tpl = c(1,2)
lst = list("firstName" = 'Britt', "lastName" = 'Anderson')
df = data.frame('fn' = c("bob","jane","griffin"),"gndr" = c('m','f','o'))
df
```

Each of three datatypes in R are demonstrated. 
:::

## Data.Frames

`Data.Frames` are particularly central to the R analysis model and are a killer feature. Think of `data.frames` as a supercharged spreadsheet with columnn names.


```{r}
is.data.frame(cars)
head(cars)
```

`cars` is a data set built in the R language. We use the `is.data.frame` command to ask if a variable is or is not a data.frame.

## Selecting Data

A bit of code that tests for whether something is true or false can be called a *predicate*. Predicates are important when having your code do something conditionally. If this is true then do this, otherwise do that. This common construct of "if-then-else" exists in most programming languages. In the next code snippet we use the conditional expressions to select on a subset of our data. You might use something similar to figure out how many of the participants in an experiment were Arts students and under 20 years of age.
    

::: {#fig-booleansinR}

```{r}
length(cars$dist[cars$speed > 10 & cars$speed < 20])

```

How many cars are there that can go faster than 10, but not more than 20?

:::

What is going on here?
1. `cars` is the name of the data frame.
2. We access a *column* of the data frame with the dollar sign notation `cars$dist`. You can see the names of the columns in a data frame with `names(cars)`.
3. We use the square brackets to *index* the column of data we are interested in. Here we do not use specific numbers, but rather we use a *boolean* to compare the values in a column and we only include the *TRUE* values. Here we ask for all the indices where the speed is greater than 10, but less than 20, and we use those indices to get the values for the `dist` column. 


## Practice

While qmd files and Quarto are great for writing documents with code in it, they are not idea for exploring in a quick and interactive way. For that use a terminal or console. In VSCode you have a menu that will create a new terminal where you can start R with the command `R`.

1. Sort (or `order`) cars by the `dist` variable.
2. Find the mean and standard deviation of the speed of the cars.
3. Are there other datasets?   
   ```{{r}} 
   library(help="datasets")
   
   ```
4. Open any of the datasets that catches your eye.
5. What are the column names?
6. How many rows?
7. What is the *comment* designator for R?
8. What is the ending extension of an R script?


## Looping in R

::: {.callout-tip}
Loops are more common in python than R. R often allows you to "map" over data, and you will not need to loop, but when you do ...

:::

`For` loops are the most common looping construct. 

```{r}
ml <- seq(1:10)
for  (m in ml) {
  print(ml)
}
```

## More Practice

1. Edit the above so that it prints the individual number each time it goes through the loop, and not the whole list.
2. Create a list of at least 8 individual characters.
   1.Make sure they are **not** in alphabetical order
   2.Print the letters one at a time.
   3.Print the letters sorted alphabetically one at a time, but *do not* overwrite your original list.
   4.Print the letters from both lists with a `format` command that says which position the letter is in.

```{r}
#| code-fold: true
#| eval: false
#| 
myName = "brittAnderson"
myList = unlist(strsplit(myName,""))

for (l in myList){
  print(l)
  }

for (l in myList[order(myList)]){
  print(l)
  }

i = 1
for (n in order(myList)){
  t  <- sprintf("The %.0fth letter of myList is: %s, but is %s in the sorted list.",i,myList[i],myList[n])
  print(t)
  i = i+1  
  }

```

## Writing a Function in R

```{r}
myadd  <- function(x,y) { # <1>
  return(x+y)             # <2>
  }
```
1. The arrow is used to assign something to a name. Other languages usually use the equals sign. The keyword `function` tells R you are creating a function, and the parentheses following establishes the number of inputs and the nicknames you will use in your function. 
2. `return` specifies what we want to get back from the function. 


Now we can use the function we defined with particular inputs. 

```{r} 
myadd(2,3)
```
